Generating Correlated Data for Omics Simulation

### 2025-05-21 manuscript/paper.qmd
Add benchmarking of vine copula method
And various minor edits in manuscript

We first discuss the simplest case, where our dataset is multivariate normally distributed.
The distribution $N(\mu, \Sigma)$ is the multivariate normal distribution with mean vector $\mu$ and covariance matrix $\Sigma$.
Here $\mu$ is a $p \times 1$ column vector and $\Sigma$ is a $p \times p$ matrix.
In order for this to work, $\Sigma$ must be symmetric and positive semi-definite, meaning that all of its eigenvalues are non-negative. 
This matrix is conceptually simple, since $\Sigma_{ij}$ gives the the covariance of $x_i$ and $x_j$ when $x \sim N(\mu, \Sigma)$.
More specifically, this is the population covariance matrix of $N(\mu, \Sigma)$, which does not generally equal the sample covariance matrix.
Indeed, if $X$ is $n$ samples from $N(\mu, \Sigma)$, then $\hat \Sigma := (X - \bar X) (X - \bar X)^T / (n - 1)$ is the sample covariance matrix where $\bar X$ is the average of the $n$ samples.
For large $n$, $\hat \Sigma$ will closely approximate $\Sigma$, but we care primarily about the situation where $n$ is small.
In particular, $\hat \Sigma$ is at most a rank $n - 1$ matrix while $\Sigma$ could be up to rank $p$.
This means that $N(\mu, \hat \Sigma)$ and $N(\mu, \Sigma)$ are quite different distributions: every sample from $N(\mu, \hat \Sigma)$ is contained in an $n-1$ dimensional plane.
If $n$ is small, then this is very unlike real data, which typically is close to $p$ dimensional.

@@

By default, CYCLOPS performs dimension reduction so that each dimension (called an "eigengene") contains at least 3% of the total variance.
We found that CYCLOPS performance depended significantly on this parameter, with the default producing good performance across all simulations.
However, when dropping CYCLOPS to require just 2% variance in each eigengene, we found that its performance depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS a).
At that setting, CYCLOPS performance is much higher in the PCA method and moderately improved in Wishart method, compared to the independent method. 
This difference is likely driven by the difference in the number of eigengenes used (@fig-CYCLOPS b), which is a measure of how much dependence is present in the dataset.
This demonstrates that the correlational structure of the transcriptome can have a major impact on performance.

@@

The first of our three methods attempts to match the top $k$ PCA components of the normalized reference dataset $Z$.
Specifically, let $u_1, \ldots, u_k$ be the left singular vectors of $Z$ with $\lambda_1, \ldots, \lambda_k$ the corresponding top $k$ signular values.
This method computes $\Sigma_{sim}$ such that $u_i^T \Sigma_{sim} u_i = \lambda_i^2$, i.e. that the variance in the direction of $u_i$ exactly matches of the reference dataset's variance in that same direction.
One solution is to use the sample covariance matrix, but that is not full rank and would match for all $i \leq n$ instead of just $i \leq k$.
Instead, we use the following:

@@

Since the distribution of the $\lambda'_i$ does not have a known analytic solution, we approximate this situation with the spiked Wishart distribution.
The rank $k-1$ Wishart distribution is the distribution of sample covariance matrices of $Z$ whose $n$ columns of $Z$ are iid $N(0, \Sigma_{sim})$.
The spiked Wishart is the special case where $\Sigma_{sim}$ has $k$ arbitrary eigenvalues and the remaining are all equal to a constant.
Note that the singular values of $Z$ are the square roots of the eigenvalues of $\widehat \Sigma_{ref}.
While our case has $\Sigma_{sim}$ non-diagonal, $\Sigma_{sim}$ may be diagonalized by orthogonal rotations due the sepctral theorem, and orthogonal rotations do not change the singular values of $Z$.
Therefore, the distribution of singular values is not affected by the assumption that $\Sigma_{sim}$ is diagonal.
Moreover, for the form $\Sigma_{sim} = D^2 + U W^2 U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
Therefore, when $D = c^2I$ for some constant $c$, we can approximate $\Sigma_{sim}$ as having $k$ arbitrary eigenvalues from $W^2$ and $n$ remaining eigenvalues all equal to $c^2$ corresponding to $D$.
This is a spiked Wishart distribution.

@@

We compared our simulated data to that of the R package SPsimSeq [@Assefa2020].
In contrast to our package, SPsimSeq uses a two-step randomization.
In short, it first fits distributions to each genes and then bins those distributions into discrete buckets.
Then, new data is generated by drawing from a multivariate normal distribution whose covariance matrix equals the sample covariance matrix of the reference dataset.
These values are used to choose which bucket each gene is drawn from.
Then, each gene's final value is drawn uniformly and independently from within its chosen bucket.
Therefore, SPsimSeq induces correlation in the first multivariate normal draw, but then injects additional independence in the second uniform step.
This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will dependend upon parameters such as the number of buckets used.
More fine-grained buckets will induce higher correlation, and in the limit of infinite buckets, it should approach our PCA method with $k$ equal to the number of samples in the reference dataset.
SPsimSeq is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options,.
In comparison, our dependentsimr package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.

>>>>

We first discuss the simplest case, where our dataset is multivariate normally distributed.
The distribution $N(\mu, \Sigma)$ is the multivariate normal distribution with mean vector $\mu$ and covariance matrix $\Sigma$.
Here $\mu$ is a $p \times 1$ column vector and $\Sigma$ is a $p \times p$ matrix.
In order for this to work, $\Sigma$ must be symmetric and positive semi-definite, meaning that all of its eigenvalues are non-negative. 
This matrix is conceptually simple, since $\Sigma_{ij}$ gives the covariance of $x_i$ and $x_j$ when $x \sim N(\mu, \Sigma)$.
More specifically, this is the population covariance matrix of $N(\mu, \Sigma)$, which does not generally equal the sample covariance matrix.
Indeed, if $X$ is $n$ samples from $N(\mu, \Sigma)$, then $\hat \Sigma := (X - \bar X) (X - \bar X)^T / (n - 1)$ is the sample covariance matrix where $\bar X$ is the average of the $n$ samples.
For large $n$, $\hat \Sigma$ will closely approximate $\Sigma$, but we care primarily about the situation where $n$ is small.
In particular, $\hat \Sigma$ is at most a rank $n - 1$ matrix while $\Sigma$ could be up to rank $p$.
This means that $N(\mu, \hat \Sigma)$ and $N(\mu, \Sigma)$ are quite different distributions: every sample from $N(\mu, \hat \Sigma)$ is contained in an $n-1$ dimensional plane.
If $n$ is small, then this is very unlike real data, which typically is close to $p$ dimensional.

@@

By default, CYCLOPS performs dimension reduction so that each dimension (called an "eigengene") contains at least 3% of the total variance.
We found that CYCLOPS performance depended significantly on this parameter, with the default producing good performance across all simulations.
However, when dropping CYCLOPS to require just 2% variance in each eigengene, we found that its performance depends significantly on the dependence structure of the simulated time series data (@fig-CYCLOPS a).
At that setting, CYCLOPS performance is much higher in the PCA method and moderately improved in Wishart method, compared to the independent method. 
This difference is likely driven by the difference in the number of eigengenes used (@fig-CYCLOPS b), which is a measure of how much dependence is present in the dataset.
This demonstrates that the correlation structure of the transcriptome can have a major impact on performance.

@@

The first of our three methods attempts to match the top $k$ PCA components of the normalized reference dataset $Z$.
Specifically, let $u_1, \ldots, u_k$ be the left singular vectors of $Z$ with $\lambda_1, \ldots, \lambda_k$ the corresponding top $k$ singular values.
This method computes $\Sigma_{sim}$ such that $u_i^T \Sigma_{sim} u_i = \lambda_i^2$, i.e. that the variance in the direction of $u_i$ exactly matches of the reference dataset's variance in that same direction.
One solution is to use the sample covariance matrix, but that is not full rank and would match for all $i \leq n$ instead of just $i \leq k$.
Instead, we use the following:

@@

Since the distribution of the $\lambda'_i$ does not have a known analytic solution, we approximate this situation with the spiked Wishart distribution.
The rank $k-1$ Wishart distribution is the distribution of sample covariance matrices of $Z$ whose $n$ columns of $Z$ are iid $N(0, \Sigma_{sim})$.
The spiked Wishart is the special case where $\Sigma_{sim}$ has $k$ arbitrary eigenvalues and the remaining are all equal to a constant.
Note that the singular values of $Z$ are the square roots of the eigenvalues of $\widehat \Sigma_{ref}.
While our case has $\Sigma_{sim}$ non-diagonal, $\Sigma_{sim}$ may be diagonalized by orthogonal rotations due the spectral theorem, and orthogonal rotations do not change the singular values of $Z$.
Therefore, the distribution of singular values is not affected by the assumption that $\Sigma_{sim}$ is diagonal.
Moreover, for the form $\Sigma_{sim} = D^2 + U W^2 U^T$ where $p$ is very large, each column of $U$ is typically very close to orthogonal to any $e_i$, a standard basis vector.
Therefore, when $D = c^2I$ for some constant $c$, we can approximate $\Sigma_{sim}$ as having $k$ arbitrary eigenvalues from $W^2$ and $n$ remaining eigenvalues all equal to $c^2$ corresponding to $D$.
This is a spiked Wishart distribution.

@@

We compared our simulated data to that of the R package SPsimSeq [@Assefa2020].
In contrast to our package, SPsimSeq uses a two-step randomization.
In short, it first fits distributions to each genes and then bins those distributions into discrete buckets.
Then, new data is generated by drawing from a multivariate normal distribution whose covariance matrix equals the sample covariance matrix of the reference dataset.
These values are used to choose which bucket each gene is drawn from.
Then, each gene's final value is drawn uniformly and independently from within its chosen bucket.
Therefore, SPsimSeq induces correlation in the first multivariate normal draw, but then injects additional independence in the second uniform step.
This two-step process makes it challenging to compare on theoretical grounds to our proposed methods since it will depended upon parameters such as the number of buckets used.
More fine-grained buckets will induce higher correlation, and in the limit of infinite buckets, it should approach our PCA method with $k$ equal to the number of samples in the reference dataset.
SPsimSeq is more specialized and full-featured for RNA-seq simulation, providing, for example, native differential expression (DE) options,.
In comparison, our dependentsimr package requires manually setting marginal expression values to inject DE, but also supports other marginal distributions for situations outside of RNA-seq.

### 2025-05-21 manuscript/supplemental/supplemental.qmd
Add benchmarking of vine copula method
And various minor edits in manuscript

@@

The R package [SPsimSeq](https://github.com/CenterForStatistics-UGent/SPsimSeq) [@Assefa2020] provides a dedicated RNA-seq and single-cell RNA-seq simulator using a Gaussian copula approach to simulate gene dependence.
We ran SPsimSeq using the following options:

@@

>>>>

### Benchmark
```{r}
#| label: fig-benchmark
#| fig-cap: "Comparison of (a) run time and (b) memory usage for our methods, SPsimSeq, and vine copula (implemented by the rvinecopulib R package). 100 samples were generated with a varying number of genes. GSE151564 dataset was randomly subsetted to have the required number of genes to be used as the reference data set. Vine copula method was only run up to 2000 genes due to the memory use required at higher gene counts."
#| fig-width: 8
#| fig-height: 4
library(tidyverse)
library(patchwork)
METHOD_ORDER <- c("real", "indep", "pca", "wishart", "corpcor", "SPsimSeq", "vinecopula")
METHOD_COLORS <- c("black", RColorBrewer::brewer.pal(length(METHOD_ORDER)-1, "Dark2"))
benchmark <- read_tsv("../../processed/benchmark/results.txt") |>
    filter(!is.na(time)) # drop the ones that aren't run completely due to time/memory constraints

g1 <- ggplot(benchmark, aes(x=n_genes, y=time, color=method)) +
    geom_line(linewidth=2) +
    labs(x = "N genes", y = "time (seconds)") + 
    scale_color_manual(values = METHOD_COLORS, breaks = METHOD_ORDER, name="method") +
    scale_x_log10() +
    scale_y_log10()
g2 <- ggplot(benchmark, aes(x=n_genes, y=max_rss, color=method)) +
    geom_line(linewidth=2) +
    labs(x = "N genes", y = "memory (MB)") +
    scale_color_manual(values = METHOD_COLORS, breaks = METHOD_ORDER, name="method") +
    scale_x_log10() +
    scale_y_log10()
g1 + g2 + plot_annotation(tag_levels="a") + plot_layout(guides = "collect")
```

@@

The R package SPsimSeq [@Assefa2020] provides a dedicated RNA-seq and single-cell RNA-seq simulator using a Gaussian copula approach to simulate gene dependence.
We ran SPsimSeq using the following options:

@@

### Running vine copula

We used the rvinecopulib package to fit and generate random data from a vine copula distribution.
This package is a R wrapper of the high-performance vinecopulib C++ package.
This package expects data normalized to the range of \[0, 1\].
Therefore, we chose not to use the real RNA-seq count data and instead generated correlated random number as input which follow a multivariate normal distribution with covariance matrix constructed with a randomly generated singular vector matrix and singular values 1, 2, ..., 100.
These values were then normalized to the interval \[0,1\] by the `pnorm` function.
A vine copula was fit using just the "guassian" family.
While rvinecopula supports multi-threaded operation, we ran using a single thread as with the other methods.
